Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device

### 2023-07-08 manuscript/manuscript.qmd
introduction makes more sense now. ZM is presented in the end of the intro.

An extensive array of research underlines the importance of sleep for
both mental and physical health[@ma2017; @meyer2022; @kpavlova2019;
@difrancesco2019]. Consequently, accurate sleep assessment methods are
essential for tracking sleep patterns, thereby enhancing our
comprehension of the sleep-health relationship. Furthermore, ensuring
high user-acceptability for these methods is essential in order to
conduct large-scale studies over prolonged periods.

While laboratory-based polysomnography is considered the gold standard
for objectively measuring sleep, its practicality in large-scale
epidemiological studies is limited due to its high cost, the need for
professional administration, and the substantial resources required for
specialized equipment[@vandewater2011]. As an alternative, diaries is
commonly used as low-cost and low-tech methods for sleep assessment in
population research. However, relying solely on diary-based methods may
introduce recall bias and other limitations[@moore2015]. A more feasible
approach in large-scale epidemiological studies is the use of
device-based measurement methods that can estimate sleep duration. This
method offers the advantage of being less burdensome for participants
and avoids potential biases associated with recall.

In this context, the Zmachine®️ Insight+ (ZM) emerges as a valuable
tool. Validated against polysomnography with favorable
results[@kaplan2014; @wang2015], the ZM provides comparable data without
the high costs or need for professional monitoring associated with
polysomnography. Furthermore, the ease of use of the ZM device makes it
compliant for free-living use[@pedersen2021]. This allows for the
analysis of multiple consecutive nights, as compared to single-night
recordings from PSG, thereby capturing the important variations in sleep
across multiple nights.

The introduction of body-worn accelerometers has provided an effective
and affordable alternative for objectively assessing sleep patterns in a
home environment over extended periods. These accelerometers collect
continuous, high-resolution data for several weeks without requiring
recharging, thus minimizing participant burden. Initial applications of
accelerometry for sleep and wake stage classification were based on
wrist movements. The original algorithm, developed in 1982 using simple
linear regression and validated with PSG [@webster1982], was later
refined in 1992 [@cole1992], leading to the widely used Cole-Kripke
model. Subsequent research on wrist-worn accelerometer data has employed
heuristic algorithms, advanced machine learning models, as well as
regression and deep learning techniques[@palotti2019; @cole1992;
@sazonov2004; @sadeh1994; @hees2015; @sundararajan2021].

Despite the well-developed field of accelerometer data analysis for
sleep detection from wrist and hip-worn devices, the same level of
advancement is not mirrored in studies utilizing thigh-worn
accelerometers. Methods for assessing sleep using wrist and hip-worn
accelerometers have greatly evolved over the years, employing an
extensive range of techniques. These include heuristic algorithms,
machine learning models, regression and deep learning techniques, all
tailored to the specific signal characteristics of wrist and hip-worn
devices [@palotti2019; @cole1992; @sazonov2004; @sadeh1994; @hees2015;
@sundararajan2021; @patterson_40_2023]. However, for thigh-worn
accelerometers, the landscape appears less mature, with only a handful
of studies investigating sleep detection algorithms. The majority of the
efforts are focused on delineating wakefulness from sleep, with
particular emphasis on the definition of 'waking time' and 'bedtime'
[@carlson2021; @inan-eroglu2021; @vanderberg2016; @winkler2016].
Furthermore, while strides have been made recently in estimating sleep
duration with these devices, with the introduction of a promising
algorithm and its comparison against PSG[@johansson_development_2023],
the field is still in its infancy when it comes to employing machine
learning techniques. Given the potential for accurate physical behavior
assessment that thigh-worn accelerometers
provide[@skotte_detection_2014; @arvidsson2019], a significant research
gap exists. Therefore, there is a pressing need for future studies to
develop techniques similar to those used for wrist and hip-worn
accelerometers, with the ultimate goal of establishing a more holistic,
accurate, and user-friendly method of sleep and physical activity
tracking.

>>>>

A vast body of research highlights the critical role of sleep in
maintaining both mental and physical health[@ma2017; @meyer2022;
@kpavlova2019; @difrancesco2019]. Consequently, accurate sleep
assessment methods are crucial for tracking sleep patterns and improving
our understanding of the sleep-health relationship. Furthermore, the
ease of use and high acceptability of these methods are essential to
facilitate large-scale, longitudinal studies.

While laboratory-based polysomnography (PSG) is typically considered the
gold standard for objective sleep measurement, its practicality in
large-scale epidemiological studies is hindered due to high costs, the
necessity for professional administration, and it is also subject to
potential rater bias[@vandewater2011; @lee2022]. As an alternative,
diaries are commonly employed as cost-effective and low-tech methods for
sleep assessment in population research. However, reliance on
diary-based methods may lead to recall bias and other
limitations[@moore2015]. A more feasible approach in large-scale
epidemiological studies is to use device-based measurement methods that
can estimate sleep duration. This approach offers the advantage of being
less burdensome for participants and eliminates potential biases
associated with recall.

Body-worn accelerometers have emerged as an effective and affordable
alternative for objectively assessing sleep patterns at home over
extended periods. These devices gather continuous, high-resolution data
for several weeks without the need for recharging, thus reducing
participant burden. Initial applications of accelerometry for sleep and
wake stage classification were based on wrist movements, starting with
an algorithm developed in 1982 and validated with PSG [@webster1982].
This model was later refined in 1992 [@cole1992], giving rise to the
widely used Cole-Kripke model. As the field advanced, an array of
techniques, including heuristic algorithms, machine learning models,
regression, and deep learning, were employed to analyze wrist-worn
accelerometer data[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
@hees2015; @sundararajan2021].

While wrist and hip-worn devices have benefited from extensive
methodological development, thigh-worn accelerometers have not seen the
same level of advancement. Existing studies mainly focus on
distinguishing sleep from wakefulness, with emphasis on defining 'waking
time' and 'bedtime' [@carlson2021; @inan-eroglu2021; @vanderberg2016;
@winkler2016]. Recent strides in estimating sleep duration using these
devices have been made, including the introduction of a promising
algorithm and its comparison against PSG[@johansson_development_2023].
Despite these advancements, the application of machine learning
techniques in this area is still relatively unexplored. Considering the
potential of thigh-worn accelerometers for accurate physical behavior
assessment[@skotte_detection_2014; @arvidsson2019], there is a
significant research gap. Therefore, future studies need to develop
techniques similar to those used for wrist and hip-worn accelerometers,
with the ultimate goal of establishing a more holistic, accurate, and
user-friendly method of sleep and physical activity tracking.

The Zmachine®️ Insight+ (ZM) emerges as a valuable tool within this
landscape. Favorably validated against PSG[@kaplan2014; @wang2015], the
ZM provides comparable data without the high costs or the need for
professional monitoring typically associated with PSG. Crucially, the ZM
facilitates multi-night analysis in free-living conditions due to its
ease of use[@pedersen2021], capturing the natural variations in sleep
patterns. This makes it advantageous over single-night PSG, particularly
as a gold standard data source in machine learning tasks, as it provides
multiple nights of measurements without inter-rater bias. Despite these
benefits, the ZM, like PSG, still poses a significant participant burden
and cost, reinforcing the need for more accessible alternatives like
accelerometers.

### 2023-07-08 manuscript/manuscript.tex
introduction makes more sense now. ZM is presented in the end of the intro.

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

An extensive array of research underlines the importance of sleep for
both mental and physical
health\citep{ma2017, meyer2022, kpavlova2019, difrancesco2019}.
Consequently, accurate sleep assessment methods are essential for
tracking sleep patterns, thereby enhancing our comprehension of the
sleep-health relationship. Furthermore, ensuring high user-acceptability
for these methods is essential in order to conduct large-scale studies
over prolonged periods.

While laboratory-based polysomnography is considered the gold standard
for objectively measuring sleep, its practicality in large-scale
epidemiological studies is limited due to its high cost, the need for
professional administration, and the substantial resources required for
specialized equipment\citep{vandewater2011}. As an alternative, diaries
is commonly used as low-cost and low-tech methods for sleep assessment
in population research. However, relying solely on diary-based methods
may introduce recall bias and other limitations\citep{moore2015}. A more
feasible approach in large-scale epidemiological studies is the use of
device-based measurement methods that can estimate sleep duration. This
method offers the advantage of being less burdensome for participants
and avoids potential biases associated with recall.

In this context, the Zmachine®️ Insight+ (ZM) emerges as a valuable
tool. Validated against polysomnography with favorable
results\citep{kaplan2014, wang2015}, the ZM provides comparable data
without the high costs or need for professional monitoring associated
with polysomnography. Furthermore, the ease of use of the ZM device
makes it compliant for free-living use\citep{pedersen2021}. This allows
for the analysis of multiple consecutive nights, as compared to
single-night recordings from PSG, thereby capturing the important
variations in sleep across multiple nights.

The introduction of body-worn accelerometers has provided an effective
and affordable alternative for objectively assessing sleep patterns in a
home environment over extended periods. These accelerometers collect
continuous, high-resolution data for several weeks without requiring
recharging, thus minimizing participant burden. Initial applications of
accelerometry for sleep and wake stage classification were based on
wrist movements. The original algorithm, developed in 1982 using simple
linear regression and validated with PSG \citep{webster1982}, was later
refined in 1992 \citep{cole1992}, leading to the widely used Cole-Kripke
model. Subsequent research on wrist-worn accelerometer data has employed
heuristic algorithms, advanced machine learning models, as well as
regression and deep learning
techniques\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021}.

Despite the well-developed field of accelerometer data analysis for
sleep detection from wrist and hip-worn devices, the same level of
advancement is not mirrored in studies utilizing thigh-worn
accelerometers. Methods for assessing sleep using wrist and hip-worn
accelerometers have greatly evolved over the years, employing an
extensive range of techniques. These include heuristic algorithms,
machine learning models, regression and deep learning techniques, all
tailored to the specific signal characteristics of wrist and hip-worn
devices
\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021, patterson_40_2023}.
However, for thigh-worn accelerometers, the landscape appears less
mature, with only a handful of studies investigating sleep detection
algorithms. The majority of the efforts are focused on delineating
wakefulness from sleep, with particular emphasis on the definition of
`waking time' and `bedtime'
\citep{carlson2021, inan-eroglu2021, vanderberg2016, winkler2016}.
Furthermore, while strides have been made recently in estimating sleep
duration with these devices, with the introduction of a promising
algorithm and its comparison against
PSG\citep{johansson_development_2023}, the field is still in its infancy
when it comes to employing machine learning techniques. Given the
potential for accurate physical behavior assessment that thigh-worn
accelerometers provide\citep{skotte_detection_2014, arvidsson2019}, a
significant research gap exists. Therefore, there is a pressing need for
future studies to develop techniques similar to those used for wrist and
hip-worn accelerometers, with the ultimate goal of establishing a more
holistic, accurate, and user-friendly method of sleep and physical
activity tracking.

>>>>

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

A vast body of research highlights the critical role of sleep in
maintaining both mental and physical
health\citep{ma2017, meyer2022, kpavlova2019, difrancesco2019}.
Consequently, accurate sleep assessment methods are crucial for tracking
sleep patterns and improving our understanding of the sleep-health
relationship. Furthermore, the ease of use and high acceptability of
these methods are essential to facilitate large-scale, longitudinal
studies.

While laboratory-based polysomnography (PSG) is typically considered the
gold standard for objective sleep measurement, its practicality in
large-scale epidemiological studies is hindered due to high costs, the
necessity for professional administration, and it is also subject to
potential rater bias\citep{vandewater2011, lee2022}. As an alternative,
diaries are commonly employed as cost-effective and low-tech methods for
sleep assessment in population research. However, reliance on
diary-based methods may lead to recall bias and other
limitations\citep{moore2015}. A more feasible approach in large-scale
epidemiological studies is to use device-based measurement methods that
can estimate sleep duration. This approach offers the advantage of being
less burdensome for participants and eliminates potential biases
associated with recall.

Body-worn accelerometers have emerged as an effective and affordable
alternative for objectively assessing sleep patterns at home over
extended periods. These devices gather continuous, high-resolution data
for several weeks without the need for recharging, thus reducing
participant burden. Initial applications of accelerometry for sleep and
wake stage classification were based on wrist movements, starting with
an algorithm developed in 1982 and validated with PSG
\citep{webster1982}. This model was later refined in 1992
\citep{cole1992}, giving rise to the widely used Cole-Kripke model. As
the field advanced, an array of techniques, including heuristic
algorithms, machine learning models, regression, and deep learning, were
employed to analyze wrist-worn accelerometer
data\citep{palotti2019, cole1992, sazonov2004, sadeh1994, hees2015, sundararajan2021}.

While wrist and hip-worn devices have benefited from extensive
methodological development, thigh-worn accelerometers have not seen the
same level of advancement. Existing studies mainly focus on
distinguishing sleep from wakefulness, with emphasis on defining `waking
time' and `bedtime'
\citep{carlson2021, inan-eroglu2021, vanderberg2016, winkler2016}.
Recent strides in estimating sleep duration using these devices have
been made, including the introduction of a promising algorithm and its
comparison against PSG\citep{johansson_development_2023}. Despite these
advancements, the application of machine learning techniques in this
area is still relatively unexplored. Considering the potential of
thigh-worn accelerometers for accurate physical behavior
assessment\citep{skotte_detection_2014, arvidsson2019}, there is a
significant research gap. Therefore, future studies need to develop
techniques similar to those used for wrist and hip-worn accelerometers,
with the ultimate goal of establishing a more holistic, accurate, and
user-friendly method of sleep and physical activity tracking.

The Zmachine®️ Insight+ (ZM) emerges as a valuable tool within this
landscape. Favorably validated against PSG\citep{kaplan2014, wang2015},
the ZM provides comparable data without the high costs or the need for
professional monitoring typically associated with PSG. Crucially, the ZM
facilitates multi-night analysis in free-living conditions due to its
ease of use\citep{pedersen2021}, capturing the natural variations in
sleep patterns. This makes it advantageous over single-night PSG,
particularly as a gold standard data source in machine learning tasks,
as it provides multiple nights of measurements without inter-rater bias.
Despite these benefits, the ZM, like PSG, still poses a significant
participant burden and cost, reinforcing the need for more accessible
alternatives like accelerometers.
